---
title: "Diel pre-examinations"
output: 
  pdf_document:
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

```{r startup, results='hide', include=FALSE}
## source necessary packages
library("mgcv")
library("ggplot2")
library("viridis")
library("extrafont")
library("grid")
library("reshape2")
library("gridExtra")
library("cowplot")

## create a theme to save linespace in plots
papertheme <- theme_bw(base_size=14, base_family = 'Arial') +
  theme(legend.position='top')

## load necessary data and make transformations
if (!file.exists('../data/private/bpbuoy2014-mod.rds')) {
  source("diel-buffalo.R")}
bdat <- readRDS('../data/private/bpbuoy2014-mod.rds')
bdat <- bdat[order(bdat$datetime),]

bdat5 <- readRDS('../data/private/bpbuoy2015-mod.rds')
bdat5 <- bdat[order(bdat5$datetime),]


bdat$cdom[bdat$cdom < 0]  <- NA
bdat$turb[log10(bdat$turb) > 2.9] <- NA

wasc <- readRDS('../data/private/WS-9-9-mod.rds')

## melt for atmospheric parameters
bair <- melt(bdat, id.vars = "datetime", measure.vars = c("windsp", "airtemp", "pressure", "dailyrain", "par1"))
bair$depth <- rep(127)
bair$depth[grep("par", bair$variable)] <- 108
bair$variable <- as.character(bair$variable)
bair$variable[grep("dsp", bair$variable)] <- "a.Wind (m/s)"
bair$variable[grep("air", bair$variable)] <- "b.Air T."
bair$variable[grep("pres", bair$variable)] <- "c.Air P.(hPa)"
bair$variable[grep("daily", bair$variable)] <- "d.Rain (mm)"
bair$variable[grep("par", bair$variable)] <- "e.PAR (air)."

## melt for temperature and clean temp data (temp3 rotten throughout)
btemp <- melt(bdat, id.vars = "datetime", measure.vars = c("temp1", "temp2", "temp4",
                                                           "temp5", "temp6"))
btemp$depth <- rep(0) #45, 77, 123, 218 + reg 82 depth
btemp$depth[grep("1", btemp$variable)] <- -82
btemp$depth[grep("2", btemp$variable)] <- -218
btemp$depth[grep("4", btemp$variable)] <- -45
btemp$depth[grep("5", btemp$variable)] <- -77
btemp$depth[grep("6", btemp$variable)] <- -123

btemp$value[which(btemp$value < 0)] <- NA
btemp$value[which(btemp$value > 23 & btemp$variable == "temp6")] <- NA # no idea why this is happening in temp6!

btemp$variable <- rep("i. Water T.")

## melt for two-depth limno parameters
blimno <- melt(bdat, id.vars = "datetime", measure.vars = c("ph1", "ph2", "ODOrel1","ODOrel2",
                                                          "bga2cell", "bga1cell"))
blimno$variable <- as.character(blimno$variable)
blimno$depth <- rep(-82)
blimno$depth[grep("2", blimno$variable)] <- -294

blimno$variable[grep("ph", blimno$variable)] <- "j. pH"
blimno$variable[grep("ODO", blimno$variable)] <- "k. O2 (%)"
blimno$variable[grep("bga", blimno$variable)] <- "l.Phyco (cell/ml)"

## one-depth limno parameters
## choose the higher par and clean
bdat$par2[bdat$par2 < 0] <- NA # deal with -100 000 values
bdat$par3[bdat$par3 < 0] <- NA

take3 <- which(bdat$par2 < bdat$par3) 
bdat$parfinal <- bdat$par2
bdat$parfinal[take3] <- bdat$par3[take3]

bdat$parfrac <- bdat$parfinal/bdat$par1 * 100
bdat$parfrac[bdat$parfrac > 100] <- NA # since dawn, dusk, night times are unreliable
bdat$parfrac[bdat$parfrac < 0] <- NA
bdat$parfrac[bdat$Hour != 12] <- NA

bones <- melt(bdat, id.vars = "datetime", measure.vars = c("chl", "co2corr", "parfrac"))
bones$depth <- rep(-82)
bones$variable <- as.character(bones$variable)
bones$depth[grep("frac", bones$variable)] <- -68
bones$depth[grep("2", bones$variable)] <- -89

bones$variable[grep("corr", bones$variable)] <- "g. CO2 (ppm)"
bones$variable[grep("chl", bones$variable)] <- "h. Chl (ug/L)"
bones$variable[grep("frac", bones$variable)] <- "f.%PAR (noon)"


## combine for megamelt and reorder if needed
megamelt <- rbind(bair, bones, btemp, blimno)
#megamelt <- within(megamelt, variable <- factor(variable, levels = sample(levels(Group))))

## melt for turbplot
bturb <- melt(bdat, id.vars = "datetime", measure.vars = c("co2corr", "cdom", "turb"))
bturb$variable <- as.character(bturb$variable)
bturb$value[bturb$variable == "turb"] <- log10(bturb$value[bturb$variable == "turb"])
bturb$variable[grep("co2", bturb$variable)] <- "a. CO2 (ppm)"
bturb$variable[grep("turb", bturb$variable)] <- "b. Turb (log10 NTU)"
bturb$variable[grep("dom", bturb$variable)] <- "c. CDOM (ug/L)"


```

```{r models, results='hide', include=FALSE}
## a lot of this code from diel-buffalo-models.R
lmnight <- with(bdat[bdat$isDay == FALSE,], lm(co2corr~pco2))
lmday <- with(bdat[bdat$isDay == TRUE,], lm(co2corr~pco2))
lmint <- with(bdat, lm(co2corr~pco2))

phmod <- gam(ph1 ~ ti(Time, bs = "cc") + ti(DOY) + ti(Time, DOY, bs = c("cc","tp")), 
          data = bdat)
co2mod <- gam(co2corr ~ ti(Time, bs = "cc") + ti(DOY) + ti(Time, DOY, bs = c("cc","tp")), 
            data = bdat)
if(!file.exists("../data/private/bp-diel-timemod2015.rds")) {
  source("../scripts/diel-buffalo-models.R")}
time4 <- readRDS("../data/private/bp-diel-timemod2014.rds")  
time5 <- readRDS("../data/private/bp-diel-timemod2015.rds")

if(!file.exists("../data/private/diel-wasc-timemod.rds")) {
  source("../scripts/wasc-diel-models.R")}
wascmod <- readRDS("../data/private/diel-wasc-timemod.rds")


```
## Buffalo Pound data

Buffalo Pound data exist from 2014, 2015, and we are waiting for any available 2016 data. We have enough supporting data to calculate pCO2 which needs doing for the diel paper in order to include Wascana as well (it's also interesting in its own right). Regrettably, none of the CO~2~ data from 2015 seem reliable, and the shallow water oxygen meter wasn't performing. So all we can get from here is pH-based. I've therefore shown no other 2015 data than a comparison between pH over the day and the sampling season, for the time being.

For 2014, we have ca 80 days of data every 15mins; For 2015, ca 145 days of data every 10mins. Sensor depths are mostly the same between the years:

*   Meteorological: Wind speed & direction, air temperature, relative humidity, pressure, daily rain (all 1.27m above surface), PAR (1.08m above surface); wind direction is unreliable since buouy changed direction itself...
* Shallow (82cm): Temperature, oxygen (% and abs), chl (rfu), phycocyanin (rfu), conductivity (uS/cm), pH, turbidity (NTU+); CDOM (ug/L) (62cm); CO2 (89cm)
* Deep (294cm): Temperature, oxygen (% and abs), chl (rfu), phycocyanin (rfu), conductivity (uS/cm), pH; CO2 (ppm) (185cm)
* 68cm: PAR (use highest PAR value between the two -- shading as day goes by)
* Thermistor chain: 45, 77, 123, 218; 318 for 2015 only

This is a brief document mostly with results, but also with some analysis. I've left my hypotheses and ideal scope of paper at the bottom of the document in order to avoid priming anyone reading this.

### Time series of all variables (2014)

The time series covered June 11^th^ to August 29^th^ and follows the most up to date data cleaning by Saskatoon. Note that corrected CO~2~ values are still not available for the last days of the time series which is being chased at the minute. In many variables, there were outliers in the form of spikes or sonde malfunction and I have taken these away. I have also removed all data during cleaning and maintenance periods. 

Meteorological variables showed both predictable diel and seasonal variation, and somewhat unpredictable patterns (Fig. \ref{mega}). For example, air temperature predictably increased through June and began to decrease again towards the end of August, with minima at night.  Highest PAR (measured in air not water) occurred in June consistent with long days and a direct position of the sun, however cloudy days with lower PAR occurred throughout. On the other hand, wind speed was variable at up to daily resolution and did not show strong seasonal trends. Air pressure dipped through the season and was at times, but not consistently, associated with rain events, reflecting the patchy nature of convective summer precipitation. 

Limnological variables were most stable in June, and most variable in August (Fig. \ref{mega}). Stratification occurred intermittently throughout the season as seen in vertical separation of temperature, oxygen and pH, most notably over three events in July. pH was maximal over July, while water temperature was highest in August. CO~2~ mirrored pH, and was above-atmospheric in June, at its lowest in July, and varied the most in August. The variability in pH and CO~2~ was not clearly related to variation in chl, which had a relatively low amplitude of variation throughout the summer until late August for which we are still waiting for corrected CO~2~ data. Generally, chl was highest in late July and the beginning of August. Note that the chl values should still be verified against routines sampling data from that summer.

<!--```{r tempplots, fig.cap="Water temperature at different depths over the study period"}
ggplot(data = btemp, aes(y=value, x=datetime, group=factor(depth), col=factor(depth))) +
  papertheme +
  geom_line() +
  scale_color_viridis(discrete = TRUE, begin= 0.1, end=0.8, 
                      option = 'viridis', direction = -1, name="Depth (cm)") +
  ylab("Temperature") + xlab("Date")

``` -->


```{r meltplot, plot-ref, fig.height=14, fig.width=9, fig.cap="Most measured variables over time, showing intermittent stratification \\label{mega}"}

ggplot(data = megamelt, aes(y=value, x=datetime, group=variable, col=depth)) +
  papertheme +
  geom_line() +
  facet_wrap('variable', scales='free_y', ncol=1, switch="y") +
  scale_color_viridis(discrete = FALSE, begin= 0.1, end=0.9, 
                      option = 'viridis', direction = 1, name="Depth (cm)") +
  theme(axis.title=element_blank(), legend.key.width=unit(2,"cm"), 
        strip.text.y = element_text(size = 8)) 
  
```

#### Potential data issues

We seem to have some interesting issues with PAR since the \% incidence of air PAR at depth undergoes a huge amount of variation depending on time of day -- it seems that the sensors are less reliable when light is low (based on a plot of \% PAR by hour; not shown) which is when water supposedly measures more PAR than air. It is likely to be a combination of errors at low light, and the relative position of the PAR sensors (if the water one is in light and the air one in shade, this could explain it). This is why, for simplicity, I chose \% PAR at noon for the plots. These data should be reliable.

The reliability of CDOM and turbidity needs checking (Fig. \ref{turb}) -- they have tripped out fairly often in 2016 so need to check everything is legit with these data in 2014. If the data are true, there might be reason to believe that the CDOM is autochthonously produced given the gradual increase over the season (I would doubt that overland flow would be the source given the dearth of heavy rain events in 2014, and them being in July???). Also -- what on earth is happening in August with both measures? Why are turbidity and CDOM sometimes inversely correlated? Is it just a matter of grain size differences and the way that NTU is measured?

Interestingly, noon \% PAR is inversely correlated with both CDOM and chl! (and cdom and chl are related!)


```{r turbplots, plot-ref, fig.cap="CDOM increases consistently over the summer season, which is not reflected in turbidity which peaks in July and late August. Is there a relationship with CO~2~? \\label{turb}", fig.width=7, fig.height=3}
turbplot <- ggplot(data=bturb, aes(y=value, x=datetime)) +
  papertheme +
  geom_point(size=0.3) +
  facet_grid(variable ~ ., scales = "free_y") +
  theme(axis.title = element_blank(), strip.text.y = element_text(size = 6))
relationplot <- ggplot(data=bdat[bdat$Hour == 12,], aes(y=parfrac, x=cdom)) +
  geom_point(size=0.5) +
  xlab("CDOM") + ylab("Noon %PAR in water")
plot_grid(turbplot, relationplot, ncol=2, rel_widths = c(1.7,1))
```

#### Questions

* Why is pH lower in August when water temperature is at its highest? 
    - My guess would be that respiration is starting to kick in, which might relate to the build-up of dead organic matter in the lake, increasing water temperatures increasing decomposition rates (Q10-type thing), and increasing mixing in the water column since stratification is less prevalent in August than July.
* Why does CO~2~ become so variable in August? 
    - I would again guess that this is because respiration is increasing which creates swings between what happens at night and what happens at day - The minima do not change that much, it's mostly the maxima that are increasing.

* How much should I pursue physical controls of limnology? 
    - e.g. should I develop a model to explain stratification events using wind speed, -direction and temperature or should I just focus on using an index for stratification to explain pH/CO~2~ (as Gavin suggested)? 

### Relationship between key variables identified in routines analysis
Routines analysis indicated that the tightest coupling is with pH and CO~2~, while oxygen was surprisingly unrelated to CO~2~. Was this the case also with high-resolution data? Did night change things?

The relationship between shallow pH and CO~2~ was strong, especially during night (Fig. \ref{phco2}). There was much more scatter during the day where CO~2~ was frequently lower than expected based on pH (e.g. routines data inicated pH ca 8.8 as being the cut-off between influx and efflux). The relationship between deep pH and CO~2~ was poor (not shown) suggesting that there is reasonable vertical separation in the water column over the season.

pCO~2~ had a stronger relationship with oxygen here than in the routines sampling (Fig. \ref{o2co2}), with simultaneous supersaturation being rare. Overall, the two were negatively correlated though simultaneous undersaturation was frequent. 


```{r relationplots, plot-ref, fig.width=5, fig.height= 4, fig.cap="The relationship between CO2 and pH is strong, but tighter at night than during the day. \\label{phco2}"}
ggplot(data = bdat, aes(y=co2corr, x=ph1, col=isDay)) +
  papertheme +
  geom_point(size=0.8) +
  scale_color_viridis(discrete = TRUE, alpha = 0.4,  begin= 0.3, end=0.8, 
                      option = 'viridis', direction = 1, name="", breaks = c(TRUE, FALSE), labels = c("Day", "Night")) +
 ylab(expression(paste(italic(p)*"CO"[2]~"(ppm)"))) +
  geom_abline(slope=0, intercept=400, lty=2, col="grey50") +
  xlab("pH") 

```


```{r dielplots, plot-ref, fig.width=5, fig.height=4, fig.cap="Supesaturation of both oxygen and carbon dioxide occurs at day and night, and the differences aren't huge \\label{o2co2}"}
lmplot <- ggplot(data=bdat, aes(y = co2corr, x = ODOrel1, col=TimeofDay)) +
  geom_point() +
  scale_color_viridis(discrete=TRUE, alpha = 0.4, option='viridis', begin= 0.3, end=0.8, direction = -1) +
  ylab(expression(paste(italic(p)*"CO"[2]~"(ppm)"))) +
  xlab(expression(paste("O"[2]~"(%)"))) +
  geom_abline(slope=0,intercept = 400, linetype='dotted') +
  geom_vline(xintercept = 100, linetype='dotted') +
  papertheme +
  theme(legend.title=element_blank(), legend.position='top')

labeldf <- data.frame(x=1.65, y=450, label="~atmospheric pCO2")

boxplot <- ggplot(bdat, aes(isDay, co2corr)) +
  theme_bw(base_size=11, base_family = 'Arial') +
  theme(legend.position='top') +
  geom_boxplot() +
  geom_hline(yintercept = 400, lty=2) +
  #geom_text(data = labeldf, aes(x=x, y=y, label=label), inherit.aes = FALSE) +
  ylab(expression(paste(italic('p')*'CO'[2]~'(ppm)'))) +
  scale_x_discrete(labels=c("Night", "Day")) +
  xlab("p < 0.001")

vp <- viewport(width = 0.33, height = 0.44, x=0.75, y=0.6)
print(lmplot)
print(boxplot, vp=vp)
```

\clearpage

### Diel amplitudes

Overall, night-time CO~2~ was not massively (but indeed significantly) higher than during the day, when pooled over the whole season (Fig. \ref{o2co2}). Diel fluctuations in CO~2~ did, however, occur, shown here by modeling CO~2~ based on day of year, and time of day (Fig. \ref{gams}). The pattern shows that the large amplitude in CO~2~ during August (Fig. \ref{mega}) is due to day-night differences, supporting a respiration-driven pathway. 

Diel amplitudes in pH were very similar between 2014 and 2015, however seasonality was different in that July in 2015 had very low pH and the overall maximum was less than in 2014; August 2015 pH was also higher (Fig. \ref{phgams}). This may all be due to differences in temperature and weather patterns over the year that would have to be investigate in more detail if it became an important aspect of the paper.

### Questions

* Should we try applying the Stets stuff with the oxygen-CO~2~ stuff? Pretty cool that both are simultaneously undersaturated at times!? (WHY is this??)
* I haven't plotted it but oxygen has pretty much the same pattern as CO~2~ pooled over the season - not much of a difference at all between day and night (night a bit lower) and I doubt it would be significant. This raises one important question -- could our close-to atmospheric pCO~2~ NOT reflect the lakes per se (in that they are, say, below or close to below atmospheric) but just wind mixing being sufficient to maintain CO~2~ close to equilibrium -- and therefore we never witness strong oversaturation. i.e. our lakes might have exactly the same reservoir of CO~2~ as boreal lakes, it's just that the wind keeps strong gradients from developing most of the time? Something to think about....
    - Still doesn't take away from the brill fact that July is completely under equilibrium, if we decide to think of that as a brill thing in terms of marketing prairie lakes for carbon sequestration.      - Perhaps we could test wind here  (say, mean night CO~2~ - mean day CO~2~ modeled against mean 24h, or day \& night wind?) and if it doesn't affect amplitude we could safely say that hardwaters are generally less amplitude-y than softwaters...?

```{r dielgams, plot-ref, fig.width=5,fig.height=4, fig.cap="Modeled CO~2~ based on time of day, and day of year, shown here for a few different periods of the sampling season. \\label{gams}"}
## for co2
N <- 200
simDOY <- c(168:170, 190:192, 228:230)
DOYgroup <- factor(rep(1:3, times=c(3,3,3)))
DOYgroup <- factor(rep(c('Jun 17-19', 'Jul 9-11', 'Aug 16-18'), times=c(3,3,3)))  
reptimes <- length(simDOY)
preddf <- data.frame(`Time` = rep(seq(min(bdat$Time, na.rm=TRUE),
                                      max(bdat$Time, na.rm=TRUE),
                                      length = N), times=reptimes),
                     `DOY` = rep(simDOY, each = N))
co2modpred <- predict(co2mod, newdata = preddf, type = "link")

DOYgroups <- rep(DOYgroup, each = N)

predicted <- cbind(preddf, co2modpred, DOYgroups)
names(predicted)[which(names(predicted)=='co2modpred')] <- 'pCO2'
predicted$DOYgroups <- factor(predicted$DOYgroups, levels = unique(predicted$DOYgroups)[c(1,2,3)])

labdatco2 <- data.frame(x = 3, y = 380, label = "Mean (atm)")

ggplot(predicted, aes(x = Time, y = pCO2, group= DOY, col=DOYgroups)) +
  theme_bw(base_size = 10) +
  scale_colour_brewer(name = "Date", type = 'qual', palette = 'Dark2', direction=1) +
  geom_line() +
  theme(legend.position="top") +
  xlab('Time of Day') + ylab(expression(paste(italic(p)*"CO"[2]~"(ppm)"))) +
  geom_abline(intercept = 398, slope = 0, linetype='dotted') +
  geom_text(data = labdatco2, aes(label = label, x = x, y = y), 
            show.legend = FALSE, inherit.aes = FALSE,  size = 3)

```

```{r phgams, plot-ref, fig.width=9,fig.height=4, fig.cap="Modeled pH based on time of day, and day of year, shown here for a few different periods of the sampling season for 2014 and 2015. \\label{phgams}"}
## predict for 2014
N <- 200
simDOY <- c(171:173, 198:200, 237:239)
DOYgroup <- factor(rep(c('Jun 20-22', 'Jul 17-19', 'Aug 25-27'), times=c(3,3,3)))  
reptimes <- length(simDOY)
preddf <- data.frame(`Time` = rep(seq(min(bdat$Time, na.rm=TRUE),
                                      max(bdat$Time, na.rm=TRUE),
                                      length = N), times=reptimes),
                     `DOY` = rep(simDOY, each = N))
mpred <- predict(time4, newdata = preddf, type = "link")

DOYgroups <- rep(DOYgroup, each = N)

predicted <- cbind(preddf, mpred, DOYgroups)
predicted$DOYgroups <- factor(predicted$DOYgroups, levels = unique(predicted$DOYgroups)[c(1,2,3)])
names(predicted)[which(names(predicted)=='mpred')] <- 'pH'

predph2014 <- predicted

## predict for 2015
N <- 200
simDOY <- c(159:161, 190:192, 237:239)
DOYgroup <- factor(rep(c('Jun 8-10', 'Jul 9-11', 'Aug 25-27'), times=c(3,3,3)))  
reptimes <- length(simDOY)
preddf <- data.frame(`Time` = rep(seq(min(bdat5$Time, na.rm=TRUE),
                                      max(bdat5$Time, na.rm=TRUE),
                                      length = N), times=reptimes),
                     `DOY` = rep(simDOY, each = N))
mpred <- predict(time5, newdata = preddf, type = "link")

DOYgroups <- rep(DOYgroup, each = N)

predicted <- cbind(preddf, mpred, DOYgroups)
names(predicted)[which(names(predicted)=='mpred')] <- 'pH'
predicted$DOYgroups <- factor(predicted$DOYgroups, levels = unique(predicted$DOYgroups)[c(1,2,3)])
predph2015 <- predicted

## for 2014 and 2015 together
predph2015$Year <- rep(2015)
predph2014$Year <- rep(2014)
predph <- rbind(predph2014, predph2015)
predph$DOYgroups <- factor(predph$DOYgroups, levels = unique(predph$DOYgroups)[c(4,1,5,2,3)])

ggplot(predph, aes(x = Time, y = pH, group= DOY, col=DOYgroups)) +
  papertheme +
  geom_line() +
  #scale_colour_brewer(name = "Date", type = 'seq', palette = 'PuBuGn', direction=1) +
  scale_color_viridis(name = "Date", option="inferno", discrete = TRUE, begin=0.2, end=0.8) +
  facet_wrap('Year') +
  geom_line() +
  xlab('Time of Day') + ylab('pH')
```

\clearpage

## Correspondence between measured and calculated CO~2~

Somewhat helpfully to our cause, calculated pCO~2~ has often been shown higher than actual pCO~2~ as seems to be the case for our Qu'Appelle data during the day (yes the slope is \> 1 but most scatter is below the line) (Fig. \ref{calc}). Therefore it is likely that we are underestimating their sequestration capacity based on our routines sampling. This is also consistent with the scatter in the pCO~2~-pH relationship (Fig. \ref{phco2}) where daytime things are a bit unpredictable and generally lower-than-expected pCO~2~ occurs at relatively low pH values. (could sensor fouling be an issue here?)

However, the night-time measurements are higher than calculations, suggesting that some process is changing the relationship between pH and pCO~2~ so that other things become more important... We are obviously using the conductivity-alkalinity relationship here to calculate pCO~2~ so it may simply be a matter of the DIC-conductivity relationship breaking down at night when more DIC (i.e. CO~2~) is released. Would this be plausible?

```{r calcmeasured, plot-ref, fig.cap="The relationship between measured and calculated CO2 has a lot of scatter but generally follows a 1:1 line during the day. At night, the slope is different. \\label{calc}"}
abdf <- data.frame(isDay = c(TRUE, FALSE), intercept=c(lmday$coefficients[1],lmnight$coefficients[1]), slope=c(lmday$coefficients[2], lmnight$coefficients[2]))

labeldf <- data.frame(x=c(250,250), y=c(1700,1600), label=c("Night Rsq = 0.92, slope = 1.28", "Day Rsq = 0.72, slope = 1.04"))

ggplot(data=bdat, aes(y = co2corr, x = pco2, col=isDay)) + 
  papertheme +
  geom_point() +
  geom_abline(slope=1, intercept=0, color='black', lty=2) +
  geom_abline(data=abdf, aes(slope=slope, intercept=intercept, col=isDay), alpha=1, show.legend = TRUE) +
  scale_color_viridis(discrete = TRUE, alpha = 0.4,  begin= 0.3, end=0.8, 
                      option = 'viridis', direction = 1, name="", breaks = c(TRUE, FALSE), labels = c("Day", "Night")) +
  ylab(expression(paste('Measured'~italic('p')*'CO'[2]~'('*mu*'atm)'))) +
  xlab(expression(paste('Calculated'~italic('p')*'CO'[2]~'('*mu*'atm)'))) +
  geom_text(data=labeldf, aes(x=x, y=y , label=label), inherit.aes = FALSE)
```

## My hypotheses and paper focus

Basically, I'd like to look at which controls operate on CO~2~ at diel scales in these lakes. Specifically, I'd again like to partition physical factors from metabolic factors where available (given complications where they interact, like low wind speeds promoting thermal stability and cyano blooms). Therefore I'd need chl and phycocyanin to be reliable because they are the only unequivocally metabolic variables I have at this resolution. Oxygen may be wind-equilibrated and not necessarily reflect algal abundance/respiration balances. Something to think about. 

Physical factors -- wind mixing, thermal stability, stratification.... basically could be encapsulated in one index as Gavin suggested. That would be tidy. Could include rain events but not sure since only a very few in the time series anyway?

Amplitudes - I hypothesised that diel amplitudes would be less in hardwater lakes than softwater lakes for a given level of production due to the pH being buffered. Like I alluded to before, this may only be testable if I can control for the evening-out effects of constant wind on concentrations (extremes may never be expected if the water is constantly turbulent). Based on the literature I've read so far (which supports my hypothesis!!!), the amplitudes between day and night here are smaller than in most other eutrophic water bodies. In many such, night shifts the balance from net autotrophic to net heterotrophic and is crucial in estimating a correct seasonal balance. Here, it seems less important. For example through July, pCO~2~ is below atmospheric equilibrium both day and night. August may hint towards more heterotrophic autumns once respiration occurs so it would be really interesting to get pH our from Wascana sonde through September!!! Fingers crossed. Last year's pH and calculated pCO~2~ indicate that later autumn may indeed start outgassing more -- but generally diel amplitudes here too are very low (Figs \ref{wasc}, \ref{wascbox}).

Questions:

  * How much should I include 2015? Just as a point of reference and more like supplementary material? Not at all?
  * How much should I include Wascana?
  
``` {r wasc plot-ref, fig.cap="Wascana also has very small diel fluctuations - seasonal differences are larger than diel fluxes \\label{wasc}"} 
ggplot(data=wasc, aes(y = pH, x = Date.Time, col=isDay)) +
  geom_point() +
  scale_color_viridis(discrete=TRUE, alpha = 0.4, option='viridis', begin= 0.3, end=0.8, 
                      direction = 1, name="", breaks = c(TRUE, FALSE), labels = c("Day", "Night")) +
  ylab("pH") +
  xlab("Date") +
  papertheme +
  theme(legend.title=element_blank(), legend.position='top')
```

``` {r amplitudes, plot-ref, fig.height = 2,fig.cap="Models of how date and time of day influence pH variations in Wascana. Overall differences between day and night are minute. \\label{wascbox}"}
labeldf <- data.frame(x=1.65, y=450, label="~atmospheric pCO2")
wascboxplot <- ggplot(wasc, aes(isDay, pCO2)) +
  theme_bw(base_size=10, base_family = 'Arial') +
  theme(legend.position='top') +
  geom_boxplot() +
  geom_hline(yintercept = 400, lty=2) +
  #geom_text(data = labeldf, aes(x=x, y=y, label=label), inherit.aes = FALSE) +
  ylab(expression(paste(italic('p')*'CO'[2]~'(ppm)'))) +
  scale_x_discrete(labels=c("Night", "Day")) +
  xlab("p < 0.05")

N <- 200
simDOY <- c(11:13, 19:21, 25:27)
DOYgroup <- factor(rep(c('Sep 11:13', 'Sep 19-21', 'Sep 25-27'), times=c(3,3,3)))  
reptimes <- length(simDOY)
preddf <- data.frame(`Time` = rep(seq(min(wasc$Time, na.rm=TRUE),
                                      max(wasc$Time, na.rm=TRUE),
                                      length = N), times=reptimes),
                     `Day` = rep(simDOY, each = N))
mpred <- predict(wascmod, newdata = preddf, type = "link")

DOYgroups <- rep(DOYgroup, each = N)

predicted <- cbind(preddf, mpred, DOYgroups)
predicted$DOYgroups <- factor(predicted$DOYgroups, levels = unique(predicted$DOYgroups)[c(1,2,3)])
names(predicted)[which(names(predicted)=='mpred')] <- 'pH'

timeplot <- ggplot(predicted, aes(x = Time, y = pH, group= Day, col=DOYgroups)) +
  theme_bw(base_size=10, base_family = 'Arial') +
  theme(legend.position='top') +
  geom_line() +
  #scale_colour_brewer(name = "Date", type = 'seq', palette = 'PuBuGn', direction=1) +
  scale_color_viridis(name = "Date", option="inferno", discrete = TRUE, begin=0.4, end=0.8) +
  geom_line() +
  xlab('Time of Day') + ylab('pH')

plot_grid(arrangeGrob(timeplot, wascboxplot, nrow = 1, widths = c(1.5,1)))
```
