---
title: "Diel stuff"
author: "Emma"
date: "July 1, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, results='hide', include=FALSE}
## source necessary packages
library("mgcv")
library("ggplot2")

## read in data
bp <- readRDS('../data/private/bpbuoy2014-mod.rds')
```

## Diel data: What we've got
    * Buffalo Pound 2014 three monthsish Jun-Sep with enough accompanying data to calculate CO2 flux
    * Wascana 2015 trial three weeks in September; enough data to estimate CO2 flux
    * incoming Wascana data for 2016 from June-Sep!!!
These are the parameters we have for Buffalo Pound:
    * wind speed and direction
    * air temp, pressure and relative humidity
    * daily rain
    * water (in addition to that for calculating CO2): PAR, cDOM, chlorophyll, O2, turbidity, some other algal index ('rfu')
And for Wascana (other than those for calculating CO2):
    * can use Regina weather station wind and pressure data
    * O2
## Diel data: Rationale
1. Diel oxygen, pH, CO2 etc data is scarce especially for hardwater lakes
    * We don't need mega awesome data sets for them to be worth publishing at this stage, cause so little is known
    * We can therefore simply *characterise* what is going on... but also test things a little bit...
2. Predictions and questions: Diel amplitude? What drives the amplitude? What drives the general pattern? (metabolism, convection, wind, lake depth...)
    * Hardwater lakes have dampened pH amplitudes because of chemical buffering and are less susceptible to have massive nightly CO2 release (--> less of a relationship between respiration and CO2 flux/pH)
    * Highest influx during the peak summer period. Less as days get shorter and temperature decreases and a build-up of summer's dead stuff to decompose (however complication in that respiration typically more sensitive to T than photosynthesis so plankton may keep trucking while respiration gets increasingly slow)
    * Drivers of seasonal cycles?
3. How to test....
    * lit search on amplitudes of diel O2 etc in different and similar systems + comparison with our data
    * model seasonal patterns in BP and Wascana and see how things change with time
    * model diel data (co2 for BP and pH for Wascana) against wind, temperature, O2, ...

## The diel cycles for Wascana and Buffalo are SURPRISINGLY different!
```{r, echo=FALSE}
ggplot(data=bp, aes(y = co2corr, x = DOY, col = isDay)) + # ifelse(Hour >= 8 & Hour <=20, 
    #   'red', 'black')
    #scale_color_identity() + # means it understands red and black in ifelse
    scale_color_manual(values=c('black', 'red')) +
    geom_point() +
    geom_vline(xintercept = as.numeric(datesDOY)) +
    ylab("CO2 (ppm)") +
    xlab("Day of year 2014 (vertical lines cleaning dates)")


```


    

