---
title: Diel models
author: Emma
header-includes:
- \usepackage{caption} 
output: 
  pdf_document:
    fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE,  tidy.opts=list(width.cutoff=40),tidy=TRUE, cache = TRUE)
```

```{r startup, results='hide', include=FALSE}
## source necessary packages & functions
library("mgcv")
library("itsadug")
library("ggplot2")
library("viridis")
library("extrafont")
library("grid")
library("reshape2")
library("gridExtra")
library("cowplot")
library("knitr")

source("../functions/geom_rug3.R")

## create ARn term function
myAR <- function(term, lag) {
  lagterm <- vector(length=length(term))
  for(i in 1:length(term)) {
    if(i <= lag) { lagterm[i] <- NA} else {
      lagterm[i] <- term[i-lag]
    }}
  lagterm
}

## set wd
setwd("../scripts/")

## create a theme to save linespace in plots
papertheme <- theme_bw(base_size=14, base_family = 'Arial') +
  theme(legend.position='top')

## read in data
if (!file.exists('../data/private/bpbuoy2014-mod.rds')) {
  source("diel-buffalo.R")} #creates both 2014 and 2015 data
bdat4 <- readRDS('../data/private/bpbuoy2014-mod.rds')
bdat4 <- bdat4[order(bdat4$datetime),]

bdat5 <- readRDS('../data/private/bpbuoy2015-mod.rds')
bdat5 <- bdat5[order(bdat5$datetime),]

if (!file.exists("../data/private/bp-stability2015.rds")) {
  source("./bp-stratification.R")
}
schmidt <- readRDS("../data/private/bp-stability.rds") # note this for 2014
bdat4 <- merge(bdat4, schmidt)

schmidt5 <- readRDS("../data/private/bp-stability2015.rds") # note this for 2014
bdat5 <- merge(bdat5, schmidt5)

## create index of potential convective mixing, based on diff in T btw air and water
bdat4$conv <- bdat4$airtemp - bdat4$temp4 # this is the topmost, 77cm one
bdat5$conv <- bdat5$airtemp - bdat5$temp4 # this is the topmost, 77cm one

## source necessary script
source("bp-gamplots.R")
```

## Issues with models describing CO2 (or pH) using time variables

I have selected to make models both with pH and CO2. This is because for 2014, we have CO2 data available (which led to the comparison of calculated and measured CO2 for a part of the paper), but for 2015, only pH. 

There are a few overarching questions regarding these models:

1. Since autocorrelation is a problem mostly for predictions, is it absolutely necessary to account for it here since my main aim is to just describe the data? (Time isn't necessarily informative since interannual variation in weather etc is so high... The main point was to display what happens through time, and be able to show that diel fluctuations are smaller than seasonal fluxes, and those in turn are smaller than interannual fluxes)
2. You mentioned a ```gp``` spline before, and a Gamma with log link -- is this something we still should pursue? How are ```scat``` and ```Gamma``` different in helping with heteroscedasticity?
    a. This comes back to what the final model should be: GAMM, BAM or GAM (with ```gp```)
3. How to construct a BAM wisely? e.g. I can choose to almost follow the data completely, which in a way could be what I want?

### Here are the candidate models:

__1. 2014__

``` {r timemodels, echo=TRUE}
 bdat <- transform(bdat, co2lag1 = myAR(bdat$co2corr, 1))
  
  ## create models; testing AR inclusion and comparing k with edf to set k (gam.check)
  ## note that "in recent mgcv versions the output in summary() is more reliable than
  ##    the generalised likelihood ratio test we might do to compare a selected model
  ##    with a null one" (GS Oct 16)
  ## FIXME: DOY is eating up as many k's as possible; 80 too many, using most of 70
  ## FIXME: update model by Gav's ryver comments! "Made some progress with time model. 
  ## Change "tp" to "gp" for both DoY terms and remove any k values. Use Gamma and log link. 
  ##    Model will be ok-ish for now. Will look at model with predictors in AM"
  
# interaction term significant -->
  mco2te <- gam(co2corr ~ te(Time, DOY, bs = c("cc","tp")) + s(co2lag1),
                data = bdat, select = TRUE, method = "REML", family = scat,
                na.action = na.exclude)
  ## co2 dependence on previous value dwarfs all other terms... not sure I'm including everything
  ##    correctly? was I meant to do the autoregression separately?
  
  plot(acf(resid(mco2te), na.action = na.exclude)) # still autocorrelated.. could add AR2?

## try gamm approach following fromthebottomoftheheap?
## FIXME: should we pursue this?
ctrl <- list(niterEM = 0, msVerbose = FALSE, optimMethod="L-BFGS-B")
testingnull <- gamm(co2corr ~te(Time, DOY, bs = c("cc","tp")), data = bdat, method="REML",
                    control = ctrl, verbosePQL = FALSE)
testing1 <- gamm(co2corr ~te(Time, DOY, bs = c("cc","tp")), data = bdat, method="REML",
                 correlation = corARMA(form = ~ 1, p=1),
                 control = ctrl, verbosePQL = FALSE)
##testing2 <- gamm(co2corr ~te(Time, DOY, bs = c("cc","tp")), data = bdat, 
       ##          correlation = corARMA(form = ~ 1, p=2),
      ##           control = ctrl, verbosePQL = TRUE)
## FIXME: "Error in `coef<-.corARMA`(`*tmp*`, value = value[parMap[, i]]) : 
##    Coefficient matrix not invertible""
```

``` {r gammresids}
res <- resid(testing1$lme, type = "normalized")
acf(res, lag.max = 200, main = "ACF - AR1 errors for gamm (2014)")

```

__2. 2015__ -- note here, much more data so bam became necessary. For ```gamm()```, I got ```## for this model to run, I had to kill firefox, if I could run it at all:  "Error in recalc.corAR1(cSt, list(Xy = as.matrix(val))) : 'Calloc' could not allocate memory (305795169 of 8 bytes)" ``` I decided to try ```bam()```

``` {r bammod, echo=TRUE}
## trying bam to see if model fitting improves cause I can perhaps make more complex
## http://www.sfs.uni-tuebingen.de/~jvanrij/Tutorial/GAMM.html#setting-up-a-gamm-model
bamnull <- bam(ph1 ~te(Time, DOY, bs = c("cc","tp")), data = bdat5)

resnull <- resid(bamnull, type = "scaled.pearson")

bamAR1 <- bam(ph1 ~te(Time, DOY, bs = c("cc","tp"), k=25), data = bdat5, rho=0.85, method='REML')
## FIXME: how to choose k intelligently? Can almost just choose to exactly follow the data
## FIXME: could play with setting rho a bit more -- obvs ACF reduces the higher I set it
```

``` {r bamresids}
## NOTE! To get the AR1-accounted residuals here, need to use function 
op <- par(mfrow=c(1,2))
acf(resnull, main = "BAM AR0 residuals (2015)", ylab=NULL)
acf_resid(bamAR1, main = "BAM AR1 residuals (2015)", ylab="avgACF")
par(op)
#https://cran.r-project.org/web/packages/itsadug/vignettes/acf.html

```

### Mechanistic models (CO2 ~ . for 2014, pH ~ . for 2014 and 2015)

These contain the model variables we all selected together in that email chain a good while ago. Here, I was mainly confused by two things

1. All variables were apparently significant, even those with a near-0 "effect". I was confused about whether the ```gam()``` was potentially mistaken....
2. Conductivity shows a very illogical relationship, suggesting it does not have a direct relationship with the ```y``` variable but may be modulated by something else (which in a way makes sense here since it is our proxy for DIC)
3. WHY is 2015 so different from 2014??

__1. CO2 ~ . (2014)__

``` {r co2mechmod, echo=TRUE}
co2mechmod <- gam(co2corr ~ s(chlrfu, by = factor(TimeofDay), k=4) + 
                   s(log(bga1rfu), by = factor(TimeofDay), k=4) + 
                   s(airtemp) + s(log(stability+1), k=3) + s(log(dailyrain+1), k=4) + s(conv) + 
                   s(log(turb), k=4) + s(cdom, k=3) + s(windsp) + s(cond1, k=4),
                 data = bdat4, select = TRUE, method = "REML", family = tw,
                 na.action = na.exclude, control = gam.control(nthreads = 3, trace = FALSE))
```

``` {r co2mechmodplots}
op <- par(mfrow = c(2,2))
gam.check(co2mechmod)
par(op)

summary(co2mechmod)

plot_grid(plotlist = plotlist, ncol=2, nrow=4)
```

__2. pH ~ . (2014, 2015)__

Model exactly the same for both years:

```{r phmechmod, echo=TRUE}
phmechmod4 <- gam(ph1 ~ s(log(chlrfu), by = factor(TimeofDay), k=4) + s(log(bga1rfu), by = factor(TimeofDay), k=4) + 
               s(airtemp) + s(log(stability +1), k=3) + s(log(dailyrain+1), k=4) + s(conv) + 
               s(log(turb), k=4) + s(cdom, k=3) + s(windsp) + s(cond1, k=4),
             data = bdat, select = TRUE, method = "REML", family = gaussian(),
             na.action = na.exclude, control = gam.control(nthreads = 3, trace = FALSE))
# had to put k for some vars to reduce overfitting

phmechmod5 <- gam(ph1 ~ s(log(chlrfu), by = factor(TimeofDay), k=4) + s(log(bga1rfu), by = factor(TimeofDay), k=4) + 
               s(airtemp) + s(log(stability +1), k=3) + s(log(dailyrain+1), k=4) + s(conv) + 
               s(log(turb), k=4) + s(cdom, k=3) + s(windsp) + s(cond1, k=4),
             data = bdat5, select = TRUE, method = "REML", family = gaussian(),
             na.action = na.exclude, control = gam.control(nthreads = 3, trace = FALSE))
```

In 2014, 91% is explained; in 2015, 70%. Some bizarre things are occurring anyway which I'm checking with Helen, like non-overlapping conductivity between 2014 and 2015. cdom looks like in different units, and so does dailyrain. Checking against our long-term data set, conductivity registered by Helen's sonde was higher in 2015 than any historically recorded conductivity by undergrads...

Here's the plots for the two years (2014, then 2015).. Note - convection seems to be happily pulling pH below recorded pH even for recorded values of convection?

``` {r phplots, plot-ref, fig.height=13, fig.width=9, fig.cap="2014 pH model"}
plot_grid(plotlist = plotlist2014, ncol=2, nrow=4)
```



```{r phplots2015, plot-ref, fig.height=13, fig.width=9, fig.cap="2015 pH model"}
plot_grid(plotlist = plotlist2015, ncol=2, nrow=4)
```

