---
title: "Diel stuff"
author: "Emma"
date: "July 1, 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warnings = FALSE)
```

```{r, echo=FALSE, results='hide', include=FALSE}
## source necessary packages
library("mgcv")
library("ggplot2",warn.conflicts = FALSE, quietly=TRUE)
library("viridis")

## read in data
bp <- readRDS('../data/private/bpbuoy2014-mod.rds')
ww <- readRDS('../data/private/WS-9-9-mod.rds')
bpmod <- readRDS('../data/private/bpmod.rds')
bpphmod <- readRDS('../data/private/bpco2phmod.rds')
```

## Diel data
1. What we've got overall:
    * Buffalo Pound 2014 three monthsish Jun-Sep with enough accompanying data to calculate CO2 flux
    * Wascana 2015 trial three weeks in September; enough data to estimate CO2 flux
    * incoming Wascana data for 2016 from June-Sep!!!
    
2. These are the parameters we have for Buffalo Pound:
    * wind speed and direction
    * air temp, pressure and relative humidity
    * daily rain
    * water (in addition to that for calculating CO2): PAR, cDOM, chlorophyll, O2, turbidity, some other algal index ('rfu')

3. And for Wascana (other than those for calculating CO2):
    * can use Regina weather station wind and pressure data
    * O2
    
## Diel data: Rationale
1. Diel oxygen, pH, CO2 etc data is scarce especially for hardwater lakes
    * We don't need mega awesome data sets for them to be worth publishing at this stage, cause so little is known
    * We can therefore simply *characterise* what is going on... but also test things a little bit...
2. Predictions and questions: Diel amplitude? What drives the amplitude? What drives the general pattern? (metabolism, convection, wind, lake depth...)
    * Hardwater lakes have dampened pH amplitudes because of chemical buffering and are less susceptible to have massive nightly CO2 release (--> less of a relationship between respiration and CO2 flux/pH)
    * Highest influx during the peak summer period. Less as days get shorter and temperature decreases and a build-up of summer's dead stuff to decompose (however complication in that respiration typically more sensitive to T than photosynthesis so plankton may keep trucking while respiration gets increasingly slow)
    * Drivers of seasonal cycles?
3. How to test....
    * lit search on amplitudes of diel O2 etc in different and similar systems + comparison with our data
    * model seasonal patterns in BP and Wascana and see how things change with time
    * model diel data (co2 for BP and pH for Wascana) against wind, temperature, O2, ...

## The diel cycles for Wascana and Buffalo are SURPRISINGLY different!

```{r, echo=FALSE,message = FALSE, warnings = FALSE}
m3 <- gam(ph1 ~ ti(Time, bs = "cc") + ti(DOY) + ti(Time, DOY, bs = c("cc","tp")), 
          data = bp)

N <- 200
simDOY <- c(171:173, 211:213, 237:239)
DOYgroup <- factor(rep(1:3, times=c(3,3,3)))
DOYgroup <- factor(rep(c('171:173', '211:213', '237:239'), times=c(3,3,3)))  
reptimes <- length(simDOY)
preddf <- data.frame(`Time` = rep(seq(min(bp$Time, na.rm=TRUE),
                                      max(bp$Time, na.rm=TRUE),
                                      length = N), times=reptimes),
                     `DOY` = rep(simDOY, each = N))
m3pred <- predict(m3, newdata = preddf, type = "link")

DOYgroups <- rep(DOYgroup, each = N)

predicted <- cbind(preddf, m3pred, DOYgroups)
names(predicted)[which(names(predicted)=='m3pred')] <- 'pH'
ggplot(predicted, aes(x = Time, y = pH, group= DOY, col=DOYgroups)) +
  theme_bw(base_size = 10) +
  geom_line() +
  scale_colour_discrete(name="Buffalo Pound: Day of Year") +
  theme(legend.position="top") +
  xlab('Time of Day') + ylab('pH') #+
  #ggtitle("Buffalo Pound")
```

```{r, echo=FALSE,message = FALSE, warnings = FALSE}
m2 <- gam(pH ~ ti(Time, bs = "cc") + ti(DOY) + ti(Time, DOY, bs = c("cc","tp")), 
                                                  data = ww)
N <- 200
simDOY <- c(254:256, 262:264, 270:272)
DOYgroup <- factor(rep(1:3, each=3))
DOYgroup <- factor(rep(c('254:256','262:264','270:272'), each=3))  
reptimes <- length(simDOY)
preddf <- data.frame(`Time` = rep(seq(min(ww$Time, na.rm=TRUE),max(ww$Time, na.rm=TRUE),
                                  length = N), times=reptimes),
           `DOY` = rep(simDOY, each = N))
m2pred <- predict(m2, newdata = preddf, type = "link")

DOYgroups <- rep(DOYgroup, each = N)

predicted <- cbind(preddf, m2pred, DOYgroups)
names(predicted)[which(names(predicted)=='m2pred')] <- 'pH'
ggplot(predicted, aes(x = Time, y = pH, group= DOY, col=DOYgroups)) +
  theme_bw(base_size = 10) +
  geom_line() +
  scale_colour_discrete(name="Wascana: Day of Year") +
  theme(legend.position="top") +
  xlab('Time of Day') + ylab('pH') #+
  #ggtitle("Wascana")

```

## Buffalo Pound does not seem to have much simultaneous supersaturation of CO2 and O2    
```{r, echo=FALSE,message = FALSE, warnings = FALSE}
dielplot <- ggplot(data=bp, aes(y = co2corr, x = ODOrel1, col=TimeofDay)) +
  geom_point() +
  scale_color_viridis(discrete=TRUE, alpha = 0.4, option='viridis', begin= 0.3, end=0.8, direction = -1) +
  #scale_color_distiller(palette = 'Greens') +
  ylab(expression(paste(italic(p)*"CO"[2]~"(ppm)"))) +
  xlab(expression(paste("O"[2]~"(%)"))) +
  geom_abline(slope=0,intercept = 400, linetype='dotted') +
  geom_vline(xintercept = 100, linetype='dotted') +
  theme_bw(base_size = 10) +
  theme(legend.title=element_blank(), legend.position='top')
suppressWarnings(print(dielplot))

```

## Buffalo Pound does have much higher correlation between O2 and pH than the routines decadal data set! --> More (direct) metabolic control?
```{r, echo=FALSE,message = FALSE, warnings = FALSE}
bp <- transform(bp, daylength = sunset-sunrise)
bp$daylength <- as.numeric(bp$daylength)

#bpmod <- gam(ph1 ~
  #                  s(chl) +
  #                  s(airtemp) +
   #                 s(daylength) +
    #                s(ODOrel1),
     #             data = bp,
      #            select = TRUE, method = "REML", family = gaussian(),
       #           na.action = na.exclude,
        #          control = gam.control(nthreads = 3, trace = TRUE,
         #                               newton = list(maxHalf = 60)))

N <- 200
varWant <- c("chl", "airtemp", "daylength")
lakeXbar <- as.data.frame(t(colMeans(bp[,varWant], na.rm = TRUE)))

ph.pdat <- data.frame(`ODOrel1` = seq(min(bp$ODOrel1, na.rm = TRUE),
                                                    max(bp$ODOrel1, na.rm = TRUE),
                                                    length = N),
                      `chl` = rep(lakeXbar$chl, times=N),
                      `airtemp` = rep(lakeXbar$airtemp, times=N),
                      `daylength` = rep(lakeXbar$daylength, times=N))
ph.pred <- predict(bpmod, newdata = ph.pdat, type = "link", se.fit = TRUE)
predicted <- cbind(ph.pred, ph.pdat)
names(predicted)[which(names(predicted)=="fit")] <- 'pH'
ggplot(predicted, aes(x = ODOrel1, y = pH)) +
  theme_bw() +
  geom_line() +
  xlab('Oxygen (%)') + ylab('pH') #+
  #ggtitle("Wascana")

```